---
title: "coziR Tutorial"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This tutorial demonstrates how to use `coziR` for neighborhood preference analysis with the bundled simulated example data.

## Loading the Data

```{r load_data}
devtools::load_all()
library(ggplot2)

# Use package tutorial source data (works when run from repo root or tutorial dir)
data_path <- if (file.exists("tutorial_data/test_sim_data.csv")) {
  "tutorial_data/test_sim_data.csv"
} else {
  "tutorial/tutorial_data/test_sim_data.csv"
}
sim_data <- read.csv(data_path)

head(sim_data)
str(sim_data)

coords <- as.matrix(sim_data[, c("x", "y")])
labels <- sim_data$ct
```

## Visualizing the Data

```{r plot_data}
ggplot(sim_data, aes(x = x, y = y, color = factor(ct))) +
  geom_point(size = 1) +
  labs(color = "Cell Type") +
  theme_minimal()
```

## Building Neighborhood Graphs

`coziR` supports three neighborhood definitions:

1. k-nearest neighbors (`knn`)
2. Radius-based (`radius`)
3. Delaunay triangulation (`delaunay`)

```{r knn_graph}
knn_adj <- knn_graph(coords, n_neighbors = 6)
knn_adj
```

## Running COZI Analysis (KNN)

```{r run_cozi_knn}
results_knn <- run_cozi(
  coords = coords,
  labels = labels,
  nbh_def = "knn",
  n_neighbors = 6,
  n_permutations = 300,
  random_state = 42
)

results_knn$cond_ratio
results_knn$zscore
```

## Dotplot Visualization (KNN)

Color encodes `zscore`; dot size encodes `con_cell_ratio`.

```{r knn_dotplot}
knn_dotplot <- cozi_dotplot(
  results_knn,
  title = "KNN neighborhood preference dotplot"
)

knn_dotplot
```

## Comparing Neighborhood Definitions

```{r compare_methods}
results_radius <- run_cozi(
  coords = coords,
  labels = labels,
  nbh_def = "radius",
  radius = 50,
  n_permutations = 100,
  random_state = 42
)

results_delaunay <- run_cozi(
  coords = coords,
  labels = labels,
  nbh_def = "delaunay",
  n_permutations = 100,
  random_state = 42
)

results_knn$zscore
results_radius$zscore
results_delaunay$zscore
```

## Advanced Usage (under active development): Fixed Type

Fix one cell type during permutations (use label name for clarity).

```{r fixed_type}
fixed_label <- as.character(sort(unique(labels))[1])

results_fixed <- run_cozi(
  coords = coords,
  labels = labels,
  nbh_def = "knn",
  n_neighbors = 6,
  n_permutations = 1000,
  random_state = 42,
  fixed_type = fixed_label
)

results_fixed$zscore
```
